Reproducible Research Assignment #1
====
Date: Sunday, September 14, 2014
----

This the assignment #1 of **Reproducible Research Assignment** for **R markdown**.


###Loading and preprocessing the data  
1. **Load the data:**
    Please download dataset: [Activity monitoring data](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip) [52K].  
    Extract the **repdata_data_activity.zip** (activity.csv) to your current R working directory, and then load tthe data
    ```{r, echo=TRUE}
    data <- read.csv("activity.csv", header=T)
    ```
  
2. Process/transform the data (if necessary) into a format suitable for your analysis  
    Currently, the transformation of the raw data is not necessary.
  
### What is mean total number of steps taken per day? 
1. Make a histogram of the total number of steps taken each day  
    - Aggregate steps by date  
        ```{r, echo=TRUE}
        sdata <- aggregate(steps~date, data=data, FUN=sum) 
        ```
  
    - Draw histogram of the total steps taken each day 
  
        ```{r, echo=TRUE}
        hist(sdata$steps, main="Histogram of Steps Taken per Day", xlab="Steps")
        box()
        ```
  
2. Calculate and report the mean and median total number of steps taken per day
    - Calculate **mean** and **median**  
        ```{r, echo=TRUE}
        mn = mean(sdata$steps)     ## mean
        mm = median(sdata$steps)   ## median
        ```
    - The **mean** is *`r format(mn,nsmall=2)`* and **median** is
    *`r format(mm,nsmall=2)`*.

### What is the average daily activity pattern?  
1. Make a time series plot of the 5-minute interval and the average steps across
all days
    - Aggregate steps by interval  
    - Add some faking date (since date is not important but time) for diagram
    purpose
    
        ```{r, echo=TRUE}
        ddata <- aggregate(steps~interval, data=data, FUN=mean)  
        
        ## fake datetime
        l = c(strptime(c("2014/9/13"),format="%Y/%m/%d"))
        for (i in 1:287) {
            l = c(l, strptime(c("2014/9/13"),format="%Y/%m/%d") + 300*i)
        }
        ddata$time <- l  
        ```
    
    - Draw diagram
    
        ```{r, echo=TRUE}
        plot(ddata$time, ddata$steps, type="l",
            main="Average Daily Activity Patter",
            xaxt="n", xlab="Time", ylab="Average Steps")
        axis.POSIXct(side=1, ddata$time, format="%H:%M")
        ```
  
2. The maximum number of steps across all the days in the dataset of 5-minute
step average
    - Get maximum number of steps from dataste
    - Retrieve the index of maximum number of steps
    - The corresponding value of interval
    - The corresponding time (HH:MM)
        ```{r, echo=TRUE}  
        max_step = max(ddata$steps)                     ## maximum number steps  
        max_idx = which.max(ddata$steps)                ## index of max steps  
        max_itv = ddata$interval[max_idx]               ## index of interval  
        max_time = format(ddata$time[max_idx],"%H:%M")  ## corresponding time  
        ```
  The **maximum number** of average steps is *`r max_step`* and the
  corresponding **time interval** is at *`r max_time`*. 
    
### Imputing missing values
1. THe number of missing values in the dataset  
    ```{r, echo=TRUE}
    miss = is.na(data$steps)
    s_miss = sum(miss)  
    ``` 
    
    **ANS:** *`r s_miss`*
    
2. Strategy for filling in missing values in the dataset  
    - Retrieve the index list of missing value in dataset
        ```{r, echo=TRUE}  
        idx_miss = grep(TRUE, miss)
        ```  
    - **Strategy:** use the step value of the previous day or the next day  
        ```{r, echo=TRUE}  
        new_data <- data    ## new dataset without NAs
        for (i in 1:s_miss) {
            if (is.na(new_data$steps[idx_miss[i]+288])) {
                new_data$steps[idx_miss[i]] = new_data$steps[idx_miss[i]-288]
            } else {
                new_data$steps[idx_miss[i]] = new_data$steps[idx_miss[i]+288]
            }
        }
        ```
3. The new dataset (*new_data*) filling all missing values is created.  
4. Analyze the new dataset
    - Aggregate steps by date  
        ```{r, echo=TRUE}  
        new_sdata <- aggregate(steps~date, data=new_data, FUN=sum)
        ```
        
    - Histogram of the total steps taken each day
        ```{r, echo=TRUE}  
        hist(new_sdata$steps, xlab="Steps",
             main="Histogram of Steps Taken per Day (dataset without NAs)")
        box()
      
        ```
        
    - Calculate **mean** and **median**  
        ```{r, echo=TRUE}
        new_mn = mean(new_sdata$steps)      ## mean
        new_mm = median(new_sdata$steps)    ## median
        ```
        
        In the new dataset, the **mean** is *`r format(new_mn,nsmall=2)`* and
        **median** is *`r format(new_mm,nsmall=2)`*.
        
    - Use histograms of orgional dataset and new dataset to compare the differences        
        ```{r, echo=TRUE}
        hist(sdata$steps, main="Histogram of Steps Taken per Day", xlab="Steps",
             col=rgb(1,0,0,0.25))
        hist(new_sdata$steps, col=rgb(0,0,1,0.25), add=TRUE)
        legend("topright", pch=15, legend=c("NAs", "No NAs"),
               col=c(rgb(1,0,0,0.25), rgb(0,0,1,0.25)))
        box()
        ```
        
        New samples (8 days) are counted in the new dataset. And these samples
        have small sum-value of steps in the day. So, the **mean** and 
        **median** values are less than values of the original datasets
        (ignore NAs).

### Differences in activity patterns between weekdays and weekends
1. Add a new factor variable in the dataset to indicate "weekday"" and "weekend"  
    - Add weekday (1: Monday, 2: Tuesday,..., 7: Sunday)
    - Convert factor with levles of "weekday" and "weekend" 
        ```{r, echo=TRUE}
        new_data$weekday <- format(strptime(new_data$date, "%Y-%m-%d"),"%u")
        wkend = new_data$weekday %in% c("6","7")
        new_data[wkend,]$weekday <- "weekend"
        new_data[!wkend,]$weekday <- "weekday"
        new_data$weekday = factor(new_data$weekday)     ## convert to factgor
        ```
  
2. The diagrams of the average 5-minute steps in weekday and weekend
    - Aggregate steps by interval for "weekday"" and "weekend"
        ```{r, echo=TRUE}
        ## aggregate for weekday daily pattern
        week_ddata <- aggregate(steps~interval,
                                data=subset(new_data, weekday=="weekday"),
                                FUN=mean)
        
        ## aggregate for weekend daily pattern
        wend_ddata <- aggregate(steps~interval, 
                                data=subset(new_data, weekday=="weekend"),
                                FUN=mean)
    ```

    - Merge datasets of different daily patterns
    - Add some faking date (since date is not important but time) for diagram
    purpose
        ```{r, echo=TRUE}
        new_ddata <- merge(week_ddata, wend_ddata,by="interval")
        new_ddata$time <- l
        ```

    - Diagrams of weekday and weekend
        ```{r, echo=TRUE}    
        ## weekday
        plot(new_ddata$time, new_ddata$steps.x, type="l", 
             main="Average Daily Activity Patter", xaxt="n", xlab="Time",
             col=rgb(1,0,0,0.5))
        
        ## weekday
        points(new_ddata$time, new_ddata$steps.y, type="l", col=rgb(0,0,1,0.5))
        
        ## axis and legend
        axis.POSIXct(side=1, new_ddata$time, format="%H:%M")
        legend("topright", pch=15, legend=c("Weekday", "Weekend"),
               col=c(rgb(1,0,0,0.5), rgb(0,0,1,0.5)))
    ```

    - Diagram of weekday and weekend with data filtering
        ```{r, echo=TRUE}  
        ## filter
        f = c(0.03125,0.15625,0.31250,0.31250,0.15625,0.03125)
        
        ## weekday
        plot(new_ddata$time, filter(new_ddata$steps.x,filter=f), type="l", 
             main="Average Daily Activity Patter (with filter)",
             xaxt="n", xlab="Time", col=rgb(1,0,0,0.5), lwd=2)

        ## weekend
        points(new_ddata$time, filter(new_ddata$steps.y,filter=f), type="l",
               col=rgb(0,0,1,0.5), lwd=2)
        
        ## axis and legend
        axis.POSIXct(side=1, new_ddata$time, format="%H:%M")               
        legend("topright", pch=15, legend=c("Weekday", "Weekend"),
               col=c(rgb(1,0,0,0.5), rgb(0,0,1,0.5)))       
        ```  
